\mytitle

### 2015-07-19 MSc_diss.tex
Continued to work on section about model development

\subsection{Tropospheric ozone}
Tropospheric ozone ($O_3$) plays an important role in the chemical and thermal balance of the atmosphere. Although it accounts only for about 10\% of the total ozone, it is a primary precursor to the formation of hydroxyl radicals ($OH$) which in association with hydrogen peroxide ($H_2O_2$) determines the ‘cleansing’ capacity of the atmosphere \citep{Prinn2003,Tarasick2008,Thompson1992}. On the other hand, being a source of the main pollutant removal agents, ozone is an air pollutant itself. Long exposure to its high concentrations causes respiratory problems in humans and damage to sensitive plant species \citep{Fowler2008}. Last but not least characteristic of tropospheric ozone is that it is an important greenhouse gas with a current estimated radiative forcing of  $0.40\pm 0.20~Wm^{–2}$ which is approximately one fifth of the $CO_2$ radiative forcing \citep{Hartmann2013,Myhre2013}. Due to these reasons a considerable interest in quantifying tropospheric ozone concentrations, budget and trends exists for more than a 100 years \citep{Becker2004}.

Unlike many other air pollutants, tropospheric ozone is not directly emitted, but produced following the oxidation of carbon monoxide ($CO$), methane ($CH_4$), and nonmethane volatile compounds (nmVOCs) in the presence of nitrogen oxides (NOx) \citep{Crutzen1973,Myhre2013}. The emissions of these precursors arise from both natural and anthropogenic sources which have increased/intensified since pre-industrial times \citep{Parrish2014,Volz1988} (Figure 1). Presumably it has led to a roughly doubling in background tropospheric ozone concentration \citep{Guicherit2000,Hartmann2013,Tarasick2008,Vingarzan2004}, although contribution from another major source, transport from the stratosphere, may also took place \citep{Fowler2008}. Climate is also believed to be a key driver of ozone production and destruction since the chemical reactions affecting ozone concentration are strongly dependent on such climatic factors as temperature, rainfall and humidity. Therefore in a changing climate tropospheric ozone is not anymore a local air quality issue, but is a global pollution problem \citep{Fowler2008}.

\begin{figure}[h]
\includegraphics[width=\linewidth]{{./pics/Isaksen2009_O3}.png}
\caption{Observed surface ozone at different Northern Hemispheric surface stations \citep{Isaksen2009}.}
\label{fig:ratio_curves}
\end{figure}

According to the International Panel on Climate Change (IPCC) Fifth Assessment report (AR5) for present (about 2000) conditions the global mean tropospheric ozone budget is approximately 331 Tg \citep{Myhre2013}. However, despite general agreement on how the drivers impact ozone concentrations at global and regional scales, its budget varies considerably between different modelling and observational studies \citep{Stevenson2006,Wild2007,Young2012}. Table 1 demonstrates that the range of the global tropospheric ozone burden estimates is approximately 60 Tg which is 18\% of the mean. In terms of ozone production the variation is even higher and account for 28\% of the corresponding value. In that regard, two questions arise: why does the tropospheric ozone budget differ between models and observations? And secondly, why does it vary to this extent, especially in terms of production? The only way to answer these questions is to perform a rigorous investigation of the factors that drive tropospheric ozone and attribute those differences to respective factors \citep{Young2012}.

\begin{table}[h] % O3 budget
\centering
\caption{Summary of tropospheric ozone global budget model and observation estimates for present (about 2000) conditions. All uncertainties quoted as 1 standard deviation (68\% confidence interval). Adapted from \citep{Myhre2013}}
\begin{tabular}{ccl}
\hline
Burden, Tg & Production, Tg yr–1 & Reference \\
\hline
\multicolumn{3}{c}{Modelling studies} \\
$337\pm23$  & $4877\pm853$ & Young et al. (2013); ACCMIP \\
$323$	    & -	           & Archibald et al. (2011) \\
$330$	    & $4876$       & Kawase et al. (2011) \\
$312$       & $4289$       & Huijnen et al. (2010) \\
$334$       & $3826$       & Zeng et al. (2010) \\
$324$       & $4870$       & Wild and Palmer (2008) \\
$314$       & -	           & Zeng et al. (2008) \\
$319$	    & $4487$	   & Wu et al. (2007) \\
$372$	    & $5042$       & Horowitz (2006) \\
$349$	    & $4384$	   & Liao et al. (2006) \\
$344\pm39$	& $5110\pm606$ & Stevenson et al. (2006); ACCENT \\
$314\pm33$	& $4465\pm514$ & Wild (2007) (post-2000 studies) \\
\hline
\multicolumn{3}{c}{Observational studies} \\
$333$   	& -	           & Fortuin and Kelder (1998) \\
$327$	    & -	           & Logan (1999) \\
$325$	    & -	           & Ziemke et al. (2011); 60S–60N \\
$319–351$	& -	           & Osterman et al. (2008); 60S–60N \\
\hline
\end{tabular}
\end{table}

@@

\subsection{Chemical mechanism}
The chemical mechanism of the model is based on recommendations from \citep{Atkinson2004} and the Master Chemical Mechanism (http://mcm.leeds.ac.uk/MCM/home.htt). On a whole it includes 90 species and 165 reactions. It is focused on representation of the chemistry related to $O_3$ production, namely:
\begin{itemize}
\item $O_3$ null cycle, HOx and NOx cycles;
\item oxidation of methane ($CH_4$) and other alkanes, namely ethane ($C_2H_6$), propane ($C_3H_8$), n-butane ($nC_4H_{10}$), i-butane ($iC_4H_{10}$), n-pentane ($nC_5H_{12}$) and i-pentane ($nC_5H_{12}$)
\item alkyl nitrate chemistry.
\end{itemize}

Photolysis rates are calculated externally using parameters from MCMv3.3.

@@

\begin{table} % model setup: O3, CO
\caption{Model setup}
\centering
\begin{tabular}{ccc}
\hline
Compound       & Initial concentration, ppb \\
\hline
$O_3$          & 40 \\
$CO$           & 100 \\
\hline
\end{tabular}
\end{table}

\begin{table} % model setup: NOx
\caption{Model setup: NOx initial concentrations (in ppb)}
\centering
\begin{tabular}{ccc}
\hline
$NO$      & $NO_2$      & NOx  ($=NO+NO_2$) \\
\hline
0.0015    & 0.0035      & 0.005 \\
0.007     & 0.018       & 0.025 \\
0.015     & 0.035       & 0.05  \\
0.03      & 0.07        & 0.1   \\
0.07      & 0.18        & 0.25  \\
0.15      & 0.35        & 0.5   \\
0.25      & 0.5         & 0.75  \\
0.3       & 0.7         & 1     \\
0.7       & 1.8         & 2.5   \\
1.5       & 3.5         & 5     \\
3         & 7           & 10    \\
\hline
\end{tabular}
\end{table}

@@

\section{Results} \label{sec:res}

>>>>

\subsection{Tropospheric ozone}
Tropospheric ozone ($O_3$) plays an important role in the chemical and thermal balance of the atmosphere. Although it accounts only for about 10\% of the total ozone, it is a primary precursor to the formation of hydroxyl radicals ($OH$) which in association with hydrogen peroxide ($H_2O_2$) determines the ‘cleansing’ capacity of the atmosphere \citep{Prinn2003,Tarasick2008,Thompson1992}. On the other hand, being a source of main pollutant removal agents, ozone is an air pollutant itself. Long exposure to its high concentrations causes respiratory problems in humans and damage to sensitive plant species \citep{Fowler2008}. Last but not least characteristic of tropospheric ozone is that it is an important greenhouse gas with a current estimated radiative forcing of  $0.40\pm 0.20~Wm^{–2}$ which is approximately one fifth of the $CO_2$ radiative forcing \citep{Hartmann2013,Myhre2013}. Due to these reasons a considerable interest in quantifying tropospheric ozone concentrations, budget and trends exists for more than a 100 years \citep{Becker2004}.

Unlike many other air pollutants, tropospheric ozone is not directly emitted, but produced following the oxidation of carbon monoxide ($CO$), methane ($CH_4$), and nonmethane volatile compounds (nmVOCs) in the presence of nitrogen oxides (NOx) \citep{Crutzen1973,Myhre2013}. The emissions of these precursors arise from both natural and anthropogenic sources which have intensified since pre-industrial times \citep{Parrish2014,Volz1988} (Figure \ref{fig:O3observations}). Presumably it has led to roughly a doubling in background tropospheric ozone concentration \citep{Guicherit2000,Hartmann2013,Tarasick2008,Vingarzan2004}, although contribution from another major source, transport from the stratosphere, may also took place \citep{Fowler2008}. Climate is also believed to be a key driver of ozone production and destruction since the chemical reactions affecting ozone concentration are strongly dependent on such climatic factors as temperature, rainfall and humidity. Therefore in a changing climate tropospheric ozone is not anymore a local air quality issue, but is a global pollution problem \citep{Fowler2008}.

\begin{figure}[h]
\includegraphics[width=\linewidth]{{./pics/Isaksen2009_O3}.png}
\caption{Observed surface ozone at different Northern Hemispheric surface stations \citep{Isaksen2009}.}
\label{fig:O3observations}
\end{figure}

According to the Intergovernmental Panel on Climate Change (IPCC) Fifth Assessment report (AR5) for present conditions (around year 2000) the global mean tropospheric ozone budget is approximately 331 Tg \citep{Myhre2013}. However, despite general agreement on how the drivers impact ozone concentrations at global and regional scales, its budget varies considerably between different modelling and observational studies \citep{Stevenson2006,Wild2007,Young2012}. Table \ref{tab:O3budget} demonstrates that the range of the global tropospheric ozone budget estimates is approximately 60 Tg which is 18\% of the mean. In terms of ozone production the variation is even higher and account for 28\% of the corresponding value. In that regard, two questions arise: why does the tropospheric ozone budget differ between models and observations? And secondly, why does it vary to this extent, especially in terms of production? The only way to answer these questions is to perform a rigorous investigation of the factors that drive tropospheric ozone and attribute those differences to respective factors \citep{Young2012}.

\begin{table}[h] % O3 budget
\centering
\caption{Summary of tropospheric ozone global budget model and observation estimates for present (about 2000) conditions. All uncertainties quoted as 1 standard deviation (68\% confidence interval). Adapted from \citep{Myhre2013}}
\label{tab:O3budget}
\begin{tabular}{ccl}
\hline
Burden, Tg & Production, Tg yr–1 & Reference \\
\hline
\multicolumn{3}{c}{Modelling studies} \\
$337\pm23$  & $4877\pm853$ & Young et al. (2013); ACCMIP \\
$323$	    & -	           & Archibald et al. (2011) \\
$330$	    & $4876$       & Kawase et al. (2011) \\
$312$       & $4289$       & Huijnen et al. (2010) \\
$334$       & $3826$       & Zeng et al. (2010) \\
$324$       & $4870$       & Wild and Palmer (2008) \\
$314$       & -	           & Zeng et al. (2008) \\
$319$	    & $4487$	   & Wu et al. (2007) \\
$372$	    & $5042$       & Horowitz (2006) \\
$349$	    & $4384$	   & Liao et al. (2006) \\
$344\pm39$	& $5110\pm606$ & Stevenson et al. (2006); ACCENT \\
$314\pm33$	& $4465\pm514$ & Wild (2007) (post-2000 studies) \\
\hline
\multicolumn{3}{c}{Observational studies} \\
$333$   	& -	           & Fortuin and Kelder (1998) \\
$327$	    & -	           & Logan (1999) \\
$325$	    & -	           & Ziemke et al. (2011); 60S–60N \\
$319–351$	& -	           & Osterman et al. (2008); 60S–60N \\
\hline
\end{tabular}
\end{table}

@@

\subsection{Chemical mechanism}
The chemical mechanism of the model is based on recommendations from \citep{Atkinson2004} and information from the Master Chemical Mechanism, MCM v3.3 \citep{Jenkin1997,Saunders2003}, via website: http://mcm.leeds.ac.uk/MCM. (On a whole it includes 90 species and 165 reactions). The mechanism is developed to represent the chemistry that governs the production and loss of ozone, hydrocarbons and alkyl nitrates. These include:
\begin{itemize}
\item the formation of ozone as a result of the photolysis of $NO_2$ followed by the reaction of ground-state oxygen atom ($O(^3P)$) with molecular oxygen and a third body, M ($N_2$ or $O_2$);
\item the loss of ozone through its photolysis and reactions with $O(^3P)$, $OH$, $HO_2$ and $NO$, among which the last is of particular interest because it gives back $NO_2$;
\item the formation of alkylperoxy radicals ($RO_2$) from oxidation of the hydrocarbons;
\item the competition of $NO$ and $HO_2$ for reaction with $HO_2$ and $RO_2$;
\item the formation of alkyl nitrates which adds up to the competition of $NO$ and $HO_2$ for reaction with the peroxy radicals;
\item the loss of alkyl nitrates through their photolysis and the reaction with $OH$;
\item the subsequent reactions of hydroperoxides and aldehydes formed from the organic reactions mentioned above.
\item oxidation of methane ($CH_4$) and other alkanes, namely ethane ($C_2H_6$), propane ($C_3H_8$), n-butane ($nC_4H_{10}$), i-butane ($iC_4H_{10}$), n-pentane ($nC_5H_{12}$) and i-pentane ($nC_5H_{12}$)
\end{itemize}

Since the most alkoxy radicals ($RO$) react rapidly with $O_2$ under atmospheric conditions and most large (>=C4) alkoxy radicals also rapidly decompose or isomerize, alkyl nitrate formation from reaction RO+NO2 = RONO2 is relatively unimportant for most organics under atmospheric conditions while observations showed that the reaction RO2+NO = RONO2 gives a higher yield.

Since one of the main mechanisms of the alkyl nitrate production (>C3!!) is the gas gas phase oxidation of alkanes by OH in the presence of NOx.

Apart from methane in the model there are other nonmethane 
It is worth mentioning that only alkanes  
Due to purely theoretical purpose of this investigation the only sink for NOx considered in the mechanism is their loss due to formation of alkyl nitrates. It means that reactions involving formation of nitric acid ($HNO_3$) and peroxyacetyl nitrates (PAN) are not considered here, which could lead to discrepancies between our modelling results and those from other research studies.

\subsection{Photolysis}
To estimate photolysis rates for a given geographical location, one must take into account the latitude and season, as well as the time of the day.

24-hour average photolysis rates are calculated externally (outside FACSIMILE program) for 1 July using the formula and parametrisations used in MCM v3.3. The formula that describes variation of photolysis rates with solar zenith angle for the core photolysis reactions was determined by the MCM's authors by using a two stream isotopic scattering model and by calculating rates for clear sky conditions at an altitude of 0.5 km on 1 July at a latitude of 45$^{\circ}$N.
has the following form:
\begin{equation} \label{eq:MCMphotolysis}
\renewcommand{\theequation}{\arabic{equation}}
J = l(cos\chi)^mexp(-nsec\chi)
\end{equation}
where $J$ - photolysis rate coefficient, $\chi$ - solar zenith angle, $l$, $m$ and $n$ - optimised parameters.
and it is a result of determination from the available data on absorption cross section and quantum data for the core reactions using two stream  Hayman.

The values of solar zenith angle are calculated using formulas from NOAA Solar Calculator (http://www.esrl.noaa.gov/gmd/grad/solcalc/).

@@

\begin{table} % model setup: O3, CO
\caption{Fixed and variable parameters used in box model runs}
\centering
\begin{tabular}{ccc}
\hline
Parameter & Initial mixing ratio & Type \\
\hline
$H_2O$    & 2\%                  & Fixed    \\
$O_2$     & 21\%     ???         & Fixed    \\
$N_2$     & 79\%        ???      & Fixed    \\
$H_2$     & 525 ppb  ???         & Variable \\
$O_3$     & 40 ppb               & Variable \\
$CO$      & 100 ppb              & Variable \\
\hline
\end{tabular}
\end{table}

\begin{table} % model setup: NOx
\caption{Model setup: NOx initial concentrations (ppb)}
\centering
\begin{tabular}{ccc}
\hline
$NO$      & $NO_2$      & NOx  ($=NO+NO_2$) \\
\hline
0.0015    & 0.0035      & 0.005 \\
0.007     & 0.018       & 0.025 \\
0.015     & 0.035       & 0.05  \\
0.03      & 0.07        & 0.1   \\
0.07      & 0.18        & 0.25  \\
0.15      & 0.35        & 0.5   \\
0.25      & 0.5         & 0.75  \\
0.3       & 0.7         & 1     \\
0.7       & 1.8         & 2.5   \\
1.5       & 3.5         & 5     \\
3         & 7           & 10    \\
\hline
\end{tabular}
\end{table}

@@

\section{Results} \label{sec:res}
Dependence of alkyl nitrate yield on carbon number (paper about n-alkanes): bigger yield of ANs from larger carbon number means that the potential for contributing to photochemical air pollution may be less for the larger (C>6) n-alkanes that for the smaller ones.
